tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-07 paper.tex
Rejig and remove template files
Switch to lstlisting following template

%\usepackage{showframe}
\usepackage{fontenc}
\graphicspath{{Fig/}}

\usepackage{minted}

@@

% line numbers
%\usepackage[mathlines, switch]{lineno}
%\usepackage[right]{lineno}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

@@

\journaltitle{Bioinformatics}
\DOI{DOI HERE}
\copyrightyear{2023}
\pubyear{2019}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

@@

\abstract{
\textbf{Summary:}
\texttt{tstrait} is a new open-source Python library that incorporates a wide
range of state-of-the-art algorithms for simulating quantitative traits based
on an ancestral recombination graph (ARG). ARG is an active field of research
in population genetics and statistical genetics due to its promising aspects in
genome-wide association study (GWAS), and this package serves as an essential
infrastructure for utilizing ARG in GWAS simulations. In this work, we provide
an overview of the simulation framework that is incorporated in
\texttt{tstrait} and highlight some recent developments in ARG.\\
\textbf{Availability and Implementation:} \texttt{tstrait} is available as an
open-source Python package. The full documentation with examples and workflow
templates is available on \url{https://tskit.dev/tstrait/docs/}, and the
development version is maintained on Github
(\url{https://github.com/tskit-dev/tstrait}).\\ \textbf{Contact:}
\href{daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}\\
}
%\abstract{Abstracts must be able to stand alone and so cannot contain %citations to
%the paper's references, equations, etc. An abstract must consist of a single
%paragraph and be concise. Because of online formatting, abstracts must appear
%as plain as possible.}
%\keywords{keyword1, Keyword2, Keyword3, Keyword4}

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

@@

The first step of quantitative trait simulation in \texttt{tstrait} is to set
the trait model distribution by using \emph{trait\_model} function, as in the
following example:
\begin{minted}{python}
model = tstrait.trait_model(
  distribution="normal", mean=0, var=1
)
\end{minted}
The \texttt{tstrait} package supports effect size simulation from five
univariate distributions and one multivariate distribution, which are taken
from commonly used statistical distributions to simulate effect sizes
\citep{gaynor2021,haller2023}. The multivariate normal distribution trait model
is used to simulate multiple traits, assuming that they are pleiotropic. The
trait model is used as an input of \emph{sim\_trait} function to simulate
effect sizes:
\begin{minted}{python}
trait_df = tstrait.sim_trait(
   ts, num_causal=5, model=model
)
\end{minted}
\emph{ts} is the tree sequence input, and the user can set the number of causal
sites that is affecting the trait. The output will be a \texttt{pandas}
dataframe \citep{pandas} that describes the simulated effect sizes.

@@

Genetic value is simulated in \texttt{tstrait} by using the effect size dataframe:
\begin{minted}{python}
genetic_result = tstrait.sim_genetic(
   ts, trait_df, alpha=0
)
\end{minted}
\texttt{tstrait} offers flexibility in genetic value simulation, as users can
define their own effect size dataframe, in addition to using the simulated
output of \emph{sim\_trait} function. This allows users to use the output from
other simulators to simulate traits in \texttt{tstrait}.

@@

Phenotype is simulated from \texttt{tstrait} by using \emph{sim\_env}:
\begin{minted}{python}
phenotype_df = tstrait.sim_env(
   genetic_result.genetic, h2=0.3, random_seed=1
)
\end{minted}
The parameters are used to simulate environmental noise from the distribution (\ref{eq:env}).

@@

Alternatively, all the above steps of the simulation can be conducted by using
\emph{sim\_phenotype}:
\begin{minted}{python}
sim_result = tstrait.sim_phenotype(
  ts=ts, num_causal=100, model=model, h2=0.3
)
\end{minted}
The user only needs to input the number of causal sites, trait model and the
narrow-sense heritability to simulate quantitative traits. By using this
function, the user does not need to modify the details of the quantitative
trait simulation.

@@

\begin{figure}[!t]%
\centering
\includegraphics[width=213pt]{figure.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=-1$. (A) Computational time with increasing number of individuals. The
whole genome dataset from \emph{Homo sapiens} with various number of
individuals and chromosome length are simulated by using \texttt{stdpopsim}
\citep{adrion2020}. For each replicate of the simulation, \texttt{tstrait} is
used to simulate one quantitative trait with 1000 causal sites. (B)
Computational time with increasing number of causal sites on two large
realistic ARGs. \texttt{tstrait} is used to simulate quantitative traits from
the simulated French Canadian dataset \citep{anderson2023} and the inferred
tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
varying number of causal sites. The French Canadian dataset is downloaded from
\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure}

>>>>

%\usepackage{showframe}
\usepackage{fontenc}

@@

% line numbers
% \usepackage[mathlines, switch]{lineno}
% \usepackage[right]{lineno}

@@

% \journaltitle{Bioinformatics}
\journaltitle{Preprint}
% \DOI{DOI HERE}
\copyrightyear{2023}
% \pubyear{2019}
% \access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

@@

\abstract{
\textbf{Summary:}
\texttt{tstrait} is a new open-source Python library that incorporates a wide
range of state-of-the-art algorithms for simulating quantitative traits based
on an ancestral recombination graph (ARG). ARG is an active field of research
in population genetics and statistical genetics due to its promising aspects in
genome-wide association study (GWAS), and this package serves as an essential
infrastructure for utilizing ARG in GWAS simulations. In this work, we provide
an overview of the simulation framework that is incorporated in
\texttt{tstrait} and highlight some recent developments in ARG.\\
\textbf{Availability and Implementation:} \texttt{tstrait} is available as an
open-source Python package. The full documentation with examples and workflow
templates is available on \url{https://tskit.dev/tstrait/docs/}, and the
development version is maintained on Github
(\url{https://github.com/tskit-dev/tstrait}).\\ \textbf{Contact:}
\href{daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}\\
}

%\abstract{Abstracts must be able to stand alone and so cannot contain %citations to
%the paper's references, equations, etc. An abstract must consist of a single
%paragraph and be concise. Because of online formatting, abstracts must appear
%as plain as possible.}
%\keywords{keyword1, Keyword2, Keyword3, Keyword4}

@@

The first step of quantitative trait simulation in \texttt{tstrait} is to set
the trait model distribution by using \emph{trait\_model} function, as in the
following example:
\begin{lstlisting}[language=Python]
model = tstrait.trait_model(
  distribution='normal', mean=0, var=1
)
\end{lstlisting}
The \texttt{tstrait} package supports effect size simulation from five
univariate distributions and one multivariate distribution, which are taken
from commonly used statistical distributions to simulate effect sizes
\citep{gaynor2021,haller2023}. The multivariate normal distribution trait model
is used to simulate multiple traits, assuming that they are pleiotropic. The
trait model is used as an input of \emph{sim\_trait} function to simulate
effect sizes:
\begin{lstlisting}[language=Python]
trait_df = tstrait.sim_trait(
   ts, num_causal=5, model=model
)
\end{lstlisting}
\emph{ts} is the tree sequence input, and the user can set the number of causal
sites that is affecting the trait. The output will be a \texttt{pandas}
dataframe \citep{pandas} that describes the simulated effect sizes.

@@

Genetic value is simulated in \texttt{tstrait} by using the effect size dataframe:
\begin{lstlisting}[language=Python]
genetic_result = tstrait.sim_genetic(
   ts, trait_df, alpha=0
)
\end{lstlisting}
\texttt{tstrait} offers flexibility in genetic value simulation, as users can
define their own effect size dataframe, in addition to using the simulated
output of \emph{sim\_trait} function. This allows users to use the output from
other simulators to simulate traits in \texttt{tstrait}.

@@

Phenotype is simulated from \texttt{tstrait} by using \emph{sim\_env}:
\begin{lstlisting}[language=Python]
phenotype_df = tstrait.sim_env(
   genetic_result.genetic, h2=0.3, random_seed=1
)
\end{lstlisting}
The parameters are used to simulate environmental noise from the distribution (\ref{eq:env}).

@@

Alternatively, all the above steps of the simulation can be conducted by using
\emph{sim\_phenotype}:
\begin{lstlisting}[language=Python]
sim_result = tstrait.sim_phenotype(
  ts, num_causal=100, model=model, h2=0.3
)
\end{lstlisting}
The user only needs to input the number of causal sites, trait model and the
narrow-sense heritability to simulate quantitative traits. By using this
function, the user does not need to modify the details of the quantitative
trait simulation.

@@

\begin{figure}[!t]%
\centering
\includegraphics[width=213pt]{figures/time-scaling.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=-1$. (A) Computational time with increasing number of individuals. The
whole genome dataset from \emph{Homo sapiens} with various number of
individuals and chromosome length are simulated by using \texttt{stdpopsim}
\citep{adrion2020}. For each replicate of the simulation, \texttt{tstrait} is
used to simulate one quantitative trait with 1000 causal sites. (B)
Computational time with increasing number of causal sites on two large
realistic ARGs. \texttt{tstrait} is used to simulate quantitative traits from
the simulated French Canadian dataset \citep{anderson2023} and the inferred
tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
varying number of causal sites. The French Canadian dataset is downloaded from
\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure}
